Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists

### 2024-11-12 paper.tex
minor corrections

\title{Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists}
\author{}
\date{\vspace{-2.5em}November 06, 2024}

@@

While Agent-Based Modelling importance in archaeological practice is steadily growing (Romanowska and Scherjon 2023) there are significant bariers to its wider adoption. One of the sectorial problems is the access to vocational training required to gain the necessary computer programming and modelling skills. A survey of practitioners (Davies and Romanowska 2018) found significant deficiencies in the provision of ABM training; 76\% of modellers were primarily self-taught with some degree of secondary support, i.e.~27\% relied on peer support, only 7\% had any sort of formal training at a workshop or summer school, while only 14\% received training as part of a degree.

A consortium of Landward Research Teoranta, Aarhus Universitet, Leiden University and Saxion University of Applied Sciences recognised this need, and sought to address the limited opportunities that archaeologists have to access high quality training through a project aiming to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.

@@

The project team developed five online tutorials. The tutorials were first designed using storyboards and later converted into interactive guided walk-throughs using NetLOGO Web (\url{https://www.netlogoweb.org/launch}) and Javascript and hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials}). During the project, initial testing was done by all authors and issues in GitHub were used to tackle bugs.

@@

After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events all of which were run by members of the project team (Table \ref{tab:events}). Recognising the difficulty of delivering practical training in an online environment during the online events several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things. Throughout the project we sought feedback from the participants and used it to improve the tutorials prior to the following events. The feedback was obtained through two surveys using Qualtrics (\url{https://www.qualtrics.com/}). The participants of the events were asked to respond to two surveys: one before the start of the event and one completed after the end of the event. The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix 1 for the questions). At the end of the event the participants were asked to answer a second survey. Questions of this survey were aimed at measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix 2 for the questions). For some events participants had to register beforehand and we were able to send the pre-workshop survey by email. At other events no registration was possible or necessary and the pre-workshop survey was shared at the start of the event. The post-workshop survey was distributed using QR-codes or links at the end of the event.

@@

Over the course of the project three groups of students worked on the project during the Smart Solutions Semester at Saxion University of Applied Sciences. This is an interdisciplinary semester in which students of at least three different study-programmes or disciplines work together on a complex problem/project (\url{https://www.saxion.edu/business-and-research/collaborate-with-saxion/smart-solutions}). Within the semester students have to come up with new ideas or solutions to help a client with a wicked problem. The backgrounds of the various students we worked with were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media \& Game Technologies, and ICT. The aim of involving interdisciplinary groups in the project was to involve students in current research projects, to enable them to learn new things and share their innovative ideas and solutions with us. The students groups were supplementary to the deliverables of the project, but their new ideas were benificial to the project. Their help during workshops was useful, but they also were engaged in developing educational material (Aalpoel et al. 2024), testing the tutorials, developing a style for the website (Jutte et al. 2024) and other materials.

@@

Each tutorials consists of a number of individual lessons, each of which guides the learner in a self-paced manner through an increasingly challenging set of exercises that build the skills and knowledge necessary to become a confident agent-based modeller.

In the first tutorial the users learn what simulations and agent-based models are and how they can help in archaeological research (Figure \ref{fig:screenshot-lesson-3}). This tutorial consists of four lessons and makes use of the Artificial Anasazi model (Axtell et al. 2002). The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific. Various concepts related to ABM, such as the definitions of model and simulation, are introduced and the participants make their first steps in NetLogo syntax. The third lesson explains how ABMs are used in archaeological research. The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode. This tutorial builds a foundation for working in NetLogo and with ABM. All activity stays within the entry level of difficulty, although some higher level concepts are lightly touched upon.

@@

In tutorial 2 the user will learn the basics of NetLogo by making their simulation on the Out of Africal dispersal of \emph{Homo sapiens} (Young and Bettinger 1995).The learner works through the basics of NetLogo syntax and learns how to set up a simulation and visualize its outcomes. This tutorial consists of nine lessons. This tutorial intends to guide the learner from the introduction to the beginner level of proficiency. The learning curve is relatively gentle. This is achieved by working with the NetLogo web interface that enables highlighting specific words and buttons and giving visual cues as to how to proceed. The tutorial introduces several most commonly used built-in functions in Netlogo (called \emph{primitives}) and introduces the general structure of all models consisting of the initialization phase and the main simulation loop. It walks the participants through the most important NetLogo settings, such as dimensions, coordinates and the origin point of the grid on which the simulation runs. The learners gain their first stripes in programming by using simple loops and functions as well as conditional statements. The next level is achieved by introducing the topic of function definition and different types of variables. Exporting data produced by the model through plots is also explained.

In the third tutorial, the learner builds a simple trade model, again based on a published simulation (Romanowska 2018). Here learners expand their NetLogo programming skills with more complexity added to such syntactic structures as loops, lists and reporters. They are introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity. This tutorial consists of seven lessons. It intends to guide the learner from the beginner to the intermediate level of proficiency. This is achieved by introducing standards of code development such as modular code, pseudocode, debugging and annotating the code. In addition, the tutorial introduces custom agent breeds, visualization with labels and reporters, plots and monitors. This takes the learner to a more advanced level, since it develops knowledge of the programming language combined with diagnosing unexpected behaviour and problem solving.

In tutorial 4 the learner works with the seminal Sugarscape model (Epstein and Axtell 1996) to further expand their agent-based modelling and NetLogo skills. The learner explores how to set up more complex interactions between agents and the environment. Furthermore, the principles of setting up good experiments and validating models is explained. This tutorial consists of eigth lessons. It intends to solidify the learner's understanding at the intermediate level of proficiency by introducing yet more complex agent-environment interaction, alternative ways to visualize the environment and modelling spatial dynamics. Creating toy landscapes is central to lesson 5. The learner dives into the topic of experiment design at a practical level but also with regards to the scientific principles, such as validation. Thus they are guided through the process of setting up experiments in NetLogo and collecting results using monitors and plots. The topic of validation of agent-based models is critical for most archaelogical applications and thus several ways to compare simulation results with the archaeological record are explored.

In the final and fifth tutorial the participant learns more about how to incorporate agent-based modelling in archaeological research. This tutorial focuses less on programming in NetLogo and more on the principle and standards of computer-based modelling. The learner learns about the different phases of the model development process. This tutorial consists of five lessons. It summarizes the material delivered in the previous lessons into one coherent framework consistent with an intermediate level of proficiency. This tutorial approaches more theoretical aspects in a practical environment. The learner revisits all phases of the model development process starting with the conceptual phase and finishing with the analysis and interpretation of results. There is a particular focus on discussing what kind of research questions are suitable for modelling and how to pick the right modelling technique. The importance of properly conceptualizing a model before starting the technical phase of the model development is highlighted as it is a common issue for less experienced modellers. The experiment design phase, including parametrisation, validation and the analysis and interpretation of the models is explained at length since it often proves challenging for students. On a practical side, the BehaviorSpace - NetLogo's experiment environment - is explained to enable the learners to batch run their models. Finally the dissemination phase of the model development is also included to ensure that the standards in model publication are widely known and the learners explore the fundamental importance of replication for advancing scientific knowledge.

@@

As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply it or had never applied it before. The respondents were also asked how they rated the available knowledge on ABM (Figure \ref{fig:available-theory}). While 64 of them had no opinion on the subject, a proportion of the respondents (28) answered that they rated the available theory on ABM as limited and only a small group (39) as sufficient or better. This clearly shows the need for more and better educational material.

@@

The majority of the respondents (153) were enthusiastic about the teaching materials, with the majority rating them as excellent or good. The teachers were even rated better than the teaching materials, with more respondents (165) giving them the \emph{excellent} score (see Figure \ref{fig:rating-teaching}).

@@

An archaeological skills passport is a tool for tracking and verifying the archaeological skills that an individual has. The first passport was created as part of \emph{The Ardnamurchan Transitions Project}, a long running investigation of the archaeology and history of the Ardnamurchan Peninsula, Western Scotland, run by the Universities of Leicester and Manchester, with Archaeology Scotland (Cobb and Croucher 2020). Since then the passport has spread around the world to countries such as Germany (Karl et al. 2019) and the United States.~While there are many different versions, in its simplest form it is a log booklet listing skills that others can verify by signing off on their level of competencies.~

@@

The ABMA tutorials are publicly hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials/}) and are citeable (Rocks-Macqueen et al. 2024). Anyone interested can download the tutorials and their source code. To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (\href{https://ABMArchaeologists.github.io/ABMA_Website/\#/}{https://ABMArchaeologists.github.io/ABMA\_Website}) and can be followed without the need to download any software. The source code of the website is also openly available (Jutte et al. 2024).

>>>>

\title{Lessons learned: creating tutorials teaching Agent-Based Modelling for Archaeologists}
\author{}
\date{\vspace{-2.5em}November 12, 2024}

@@

While Agent-Based Modelling importance in archaeological practice is steadily growing (Romanowska and Scherjon 2023) there are significant barriers to its wider adoption. One of the problems is the access to vocational training required to gain the necessary computer programming and modelling skills. A survey of practitioners (Davies and Romanowska 2018) found significant deficiencies in the provision of ABM training; 76\% of modellers were primarily self-taught with some degree of secondary support, i.e.~27\% relied on peer support, only 7\% had any sort of formal training at a workshop or summer school, while only 14\% received training as part of a degree.

A consortium of Landward Research Teoranta, Aarhus University, Leiden University and Saxion University of Applied Sciences recognised this need, and sought to address the limited opportunities that archaeologists have to access high quality training. A project was set up to create and make available open learning materials that could be integrated into a wide range of training programmes, such as face-to-face training, employer-focussed continuing professional development seminars, webinars, MOOCs (Massive Open Online Courses) and self-directed learning.

@@

The project team developed five online tutorials. The tutorials were first designed using storyboards and later converted into interactive guided walk-throughs using NetLOGO Web (\url{https://www.netlogoweb.org/launch}) and Javascript and hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials}). During the project, initial testing was done by all authors and the issues functionality in GitHub was used to tackle bugs.

@@

After initial debugging by the project team, the tutorials were tested on a wider sample of participants during several online and in-person events all of which were run by members of the project team (Table \ref{tab:events}). Recognising the difficulty of delivering practical training in an online environment (Microsoft Teams and Zoom) during the online events several members of the team were present which enabled participants to go to break-out rooms with an expert if they needed additional help or wanted to discuss things. Throughout the project we sought feedback from the participants and used it to improve the tutorials prior to the following events. The feedback was obtained through two surveys using Qualtrics (\url{https://www.qualtrics.com/}). The participants of the events were asked to respond to two surveys: one before the start of the event and one after the end of the event. The first survey focused on the background of the participants and their original level of knowledge in relation to ABM (see appendix 1 for the questions). At the end of the event the participants were asked to answer a second survey. Questions of this survey were aimed at measuring the effectiveness of the tutorials and getting feedback on the workshop and tutorials (see appendix 2 for the questions). For some events participants had to register beforehand and we were able to send the pre-workshop survey by email. At other events no registration was possible or necessary and the pre-workshop survey was shared at the start of the event. The post-workshop survey was distributed using QR-codes or links at the end of the event.

@@

Over the course of the project three groups of students worked on the project during the Smart Solutions Semester at Saxion University of Applied Sciences. This is an interdisciplinary semester in which students of at least three different study-programmes or disciplines work together on a complex problem/project (\url{https://www.saxion.edu/business-and-research/collaborate-with-saxion/smart-solutions}). Within the semester students have to come up with new ideas or solutions to help a client with a wicked, or at least challenging, problem. The backgrounds of the various students we worked with were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media \& Game Technologies, and ICT. The aim of involving interdisciplinary groups in the project was to involve students in current research projects, to enable them to learn new things and share their innovative ideas and solutions with us. The students groups were supplementary to the deliverables of the project, but their new ideas were beneficial to the project. Their help during workshops was useful, but they also were engaged in developing educational material (Aalpoel et al. 2024), testing the tutorials, developing a style for the website (Jutte et al. 2024) and other materials.

@@

Each tutorials consists of a number of individual lessons, which guide the learner in a self-paced manner through an increasingly challenging set of exercises. Collectively, they build the skills and knowledge necessary to become a confident agent-based modeller.

In the first tutorial the users learn what simulations and agent-based models are and how they can help in archaeological research (Figure \ref{fig:screenshot-lesson-3}). This tutorial consists of four lessons and makes use of the Artificial Anasazi model (Axtell et al. 2002). The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific. Various concepts related to ABM, such as the definitions of model and simulation, are introduced and the participants take their first steps in NetLogo syntax. The third lesson explains how ABMs are used in archaeological research. The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode. This tutorial builds a foundation for working in NetLogo and with ABM. All activity stays within the entry level of difficulty, although some higher level concepts are lightly touched upon.

@@

In tutorial 2 the user will learn the basics of NetLogo by making their simulation on the Out of Africal dispersal of \emph{Homo sapiens} (Young and Bettinger 1995). The learner works through the basics of NetLogo syntax and learns how to set up a simulation and visualize its outcomes. This tutorial consists of nine lessons. This tutorial intends to guide the learner from the introduction to the beginner level of proficiency. The learning curve is relatively gentle. This is achieved by working with the NetLogo web interface that enables highlighting specific words and buttons and giving visual cues as to how to proceed. The tutorial introduces several commonly used built-in functions in Netlogo (called \emph{primitives}) and introduces the general structure of all models consisting of the initialization phase and the main simulation loop. It walks the participants through the most important features of the NetLogo world, such as dimensions, coordinates and the origin point of the grid on which the simulation runs. The learners gain their first stripes in programming by using simple loops and functions as well as conditional statements. The next level is achieved by introducing the topic of function definition and different types of variables. Exporting data produced by the model through plots is also explained.

In the third tutorial, the learner builds a simple trade model, again based on a published simulation (Romanowska 2018). Here learners expand their NetLogo programming skills with more complex syntactic structures as loops, lists and reporters. They are introduced to some key computational techniques like modular code development and debugging which become important alongside this increased coding complexity. This tutorial consists of seven lessons. It intends to guide the learner from the beginner to the intermediate level of proficiency. This is achieved by introducing standards of code development such as modular code, pseudocode, debugging and annotating the code. In addition, the tutorial introduces custom agent breeds, visualization with labels and reporters, plots and monitors. This takes the learner to a more advanced level, since it develops knowledge of the programming language combined with diagnosing unexpected behaviour and problem solving.

In tutorial 4 the learner works with the seminal Sugarscape model (Epstein and Axtell 1996) to further expand their agent-based modelling and NetLogo skills. The learner explores how to set up more complex interactions between agents and the environment. Furthermore, the principles of setting up good experiments and validating models is explained. This tutorial consists of eight lessons. It intends to solidify the learner's understanding at the intermediate level of proficiency by introducing yet more complex agent-environment interaction, alternative ways to visualize the environment and modelling spatial dynamics. Creating toy landscapes is central to lesson 5. The learner dives into the topic of experiment design at a practical level but also with regards to the scientific principles, such as validation. Thus, they are guided through the process of setting up experiments in NetLogo and collecting results using monitors and plots. The topic of validation of agent-based models is critical for most archaeological applications and thus several ways to compare simulation results with the archaeological record are explored.

In the final and fifth tutorial the learner learns more about how to incorporate agent-based modelling in archaeological research. This tutorial focuses less on programming in NetLogo and more on the principle and standards of computer-based modelling. The learner learns about the different phases of the model development process. This tutorial consists of five lessons. It summarizes the material delivered in the previous lessons into one coherent framework consistent with an intermediate level of proficiency. This tutorial approaches more theoretical aspects in a practical environment. The learner revisits all phases of the model development process starting with the conceptual phase and finishing with the analysis and interpretation of results. There is a particular focus on discussing what kind of research questions are suitable for modelling and how to pick the right modelling technique. The importance of properly conceptualizing a model before starting the technical phase of the model development is highlighted as it is a common issue for less experienced modellers. The experiment design phase, including parametrisation, validation and the analysis and interpretation of the models is explained at length since it often proves challenging for students. On a practical side, the BehaviorSpace - NetLogo's experiment environment - is explained to enable the learners to batch run their models. Finally the dissemination phase of the model development is also included to ensure that the standards in model publication are widely known and the learners explore the fundamental importance of replication for advancing scientific knowledge.

@@

As shown above, the respondents had some knowledge on ABM in general, but did not know how to apply it or had never applied it before. The respondents were also asked how they rated the available resources on ABM (Figure \ref{fig:available-theory}). While 64 of them had no opinion on the subject, a proportion of the respondents (28) answered that they rated the available theory on ABM as limited and only a small group (39) as sufficient or better. This clearly shows the need for more and better educational material.

@@

The majority of the respondents (153) were enthusiastic about the teaching materials, with the majority rating them as excellent or good. The teachers were rated even better than the teaching materials, with more respondents (165) giving them the \emph{excellent} score (see Figure \ref{fig:rating-teaching}).

@@

An archaeological skills passport is a tool for tracking and verifying the archaeological skills that an individual has. The first passport was created as part of \emph{The Ardnamurchan Transitions Project}, a long running investigation of the archaeology and history of the Ardnamurchan Peninsula, Western Scotland, run by the Universities of Leicester and Manchester, with Archaeology Scotland (Cobb and Croucher 2020). Since then, the passport has spread around the world to countries such as Germany (Karl et al. 2019) and the United States.~While there are many different versions, in its simplest form it is a log booklet listing skills that others can verify by signing off on their level of competencies.~

@@

The ABMA tutorials are publicly hosted on GitHub (\url{https://github.com/ABMArchaeologists/ABMA_tutorials/}) and are citeable (Rocks-Macqueen et al. 2024). Anyone interested can download the tutorials and their source code. To increase the accessibility and visibility of the tutorials, they are also hosted on a static website (\url{https://abmarchaeologists.github.io/ABMA_website}) and can be followed without the need to download any software. The source code of the website is also openly available (Jutte et al. 2024).
